### 2018-09-10 doc/intro.tex
expanding results section to try and explain variance sources

Measuring this pattern is extremely difficult for multiple reasons. Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).

Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound. A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0. PERFECT PLACE FOR A SIMULATION.

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/range_time}
  \caption{Geographic range over time for a selection of eight random species included in this analysis. Geographic range is measured as the maximum geodesic distance between two occurrences of that species during a given time interval. The time axis corresponds to species age in millions of years (My). Because each species has a different duration, not all of these trajectories are of equivalent duration.}
  \label{fig:range_real}
\end{figure}

Measuring this pattern is extremely difficult for multiple reasons. Species are not observed continuously through time, species geographic ranges are based on what samples are available, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).

Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound. A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0. This can be demonstrated through a simple simulation PERFECT PLACE FOR A SIMULATION.

Our simulations are of a random walk with a lower bound of 0. Each individual simulation begins with a value of 0, and for each time step that value is modified by a single realization from a standard Normal distribution. Each simulation had a maximum of 100 time steps beyond the initial point. For each simulation, the first time the value is less than or equal to 0 is considered the stopping point, with that last value set to 0. We retained all simulations which returned to 0, which would be the equivalent of a species going extinct. While we attempted 1000 simulations, not all of these fulfilled this criteria as they did not had a value greater than 0 at their final time step. Finally, the duration of all simulations was rescaled to be between 0 and 1 so that their trajectories are directly comparable. A binned average and moving-window average are calculated from the population of simulations.

The binned average and moving-window averages are nearly identical to figure 1 from the original rise-and-fall paper by Foote et al. The results from these simulations show that for a random walk bounded at 0, the average trajectory is symmetrical and virtually identical to either the ``hat'' or ``hump'' models and was generated without any deterministic process; the pattern is unimodal and intermediate between ``strict'' hat or hump models. Thus, a hat or hump trajectory demonstrates nothing about process or determinism but could just result from entirely stochastic processes.

 \begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/georange_sim_norm}
  \caption{Simulation of a bounded random walk. The blue line is a binned average of all simulations while the red line is a moving-window average of those averages.}
  \label{fig:range_sim}
\end{figure}

Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range. This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction. For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?

### 2018-09-10 doc/results.tex
expanding results section to try and explain variance sources

Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}). Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points. However, the actual differences between these models are minor as revealed by the wide the standard errors. These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
\begin{table}[ht]
  \centering
  \caption{Model comparison using WAIC and LOOIC}
  \begin{tabular}{ r r r r r }
    \hline
    Model & looic & looic se & waic & waic se \\
    \hline
    Past and vary & 12811.58 & 179.27 & 12807.61 & 170.21 \\
    No past but vary & 12834.53 & 179.07 & 12831.51 & 179.02 \\
    Past but no vary & 12836.19 & 179.20 & 12833.13 & 179.15 \\
    No past or vary & 12849.43 & 179.46 & 12847.08 & 179.43 \\
    \hline
  \end{tabular}
  \label{tab:selection}
\end{table}

@@

% variance components
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/variance_components}
  \caption{Contribution of multi-level components to unmodeled variance. Larger values indicate a greater contribution to overall variance. Variance components labeled as ``overall'' represent variance between grouping factors, while those labeled  ``within groups'' correspond to variance within grouping factors. If the overall component is greater than the within component, then the groupings do not structure the overall variance. If the within component is greater than the overall component, then the overall variance is structured by the groupings.}
  \label{fig:variance_components}
\end{figure}

% risk estimate compared to change in geo-range

>>>>

Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}). Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points. However, the actual differences between these models are minor as revealed by the wide the standard errors. These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
\begin{table}[ht]
  \centering
  \caption{Model comparison using WAIC and LOOIC}
  \begin{tabular}{ r r r r r }
    \hline
    Model & looic & looic se & waic & waic se \\
    \hline
    Past and vary & 12789.81 & 178.90 & 12785.51 & 178.84 \\
    No past but vary & 12820.67 & 178.82 & 12817.49 & 178.77 \\ 
    Past but no vary & 12847.18 & 179.40 & 12844.88 & 179.36 \\ 
    No past or vary & 12851.43 & 179.48 & 12849.10 & 179.44 \\ 
    \hline
  \end{tabular}
  \label{tab:selection}
\end{table}

@@

% effect of age at observation

% average hazard

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_baseline}
  \caption{<+caption text+>}
  \label{fig:hazard_baseline}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/hazard_bygroup}
  \caption{<+caption text+>}
  \label{fig:hazard_bygroup}
\end{figure}

% variance components
There are multiple levels and sources of variance in our model. Each of the grouping factors account for some amount of variance in the data. The first ``level'' of variance is the data level or the variance in the observed data (the model response). In the context of regular linear regression, this variance is equivalent to the residual variance. In the context of logistic regression, however, this value is not an inherent part of the model.

There are then two non-nested sources of the second ``level'' of variance: intercept and slopes varying by time of observation, and a mean-zero age at observation varying intercept. Each of the elements of the vector of scale terms \(\tau_{B}\) describe the within-phyla variance for the intercept and the slope terms for the covariates. Similarly, the scale term \(\sigma_{a}\) is a scalar that describes the within-phyla variance of the effect of age at observation.

The final ``level'' of variance the phyla-level: the between phyla variance for the intercepts and regression coefficients \(\tau_{\alpha}\), and the between phyla variance for the effect of age at observation \(\sigma_{\delta}\).

The greatest source of variance accounted for by the grouping factors is the between phylum variance in the model intercept (Fig. \ref{fig:variance_components}); this is the first element of \(\tau_{B}\). All of the other scale terms have approximately equal estimates, which means that they contribute a similar amount to the structure underlying the data. For example, the within-phyla effect of geographic range on extinction probability is associated with approximately as much variance as there is between the phyla themselves.

This result means that, given this model, the greatest contributor to changes in extinction risk over time is when a species is present. In comparison, the covariates we studied individually contribute only a limited amount to the variation in extinction. Similarly, while age at observation does contribute to some differences in extinction risk (Fig XXX age effect), this effect is most likely overwhelmed by the importance of when that observation takes place.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/variance_components}
  \caption{Contribution of multi-level components to unmodeled variance. Larger values indicate a greater contribution to overall variance. Variance components labeled as ``overall'' represent variance between grouping factors, while those labeled  ``within groups'' correspond to variance within grouping factors. If the overall component is greater than the within component, then the groupings do not structure the overall variance. If the within component is greater than the overall component, then the overall variance is structured by the groupings.}
  \label{fig:variance_components}
\end{figure}

% risk estimate compared to change in geo-range
\begin{figure}[ht]
  \centering
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_range}
    \caption{A}
    \label{fig:relrisk_range}
  \end{subfigure}

  \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/relrisk_ext}
    \caption{B}
    \label{fig:relrisk_ext}
  \end{subfigure}
  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
  \label{fig:relrisk_compare}
\end{figure}
